$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-02-23 qdpm_paper.tex
Revised sections IV & V

The QDM and QFPLM on the square lattice have isolated liquid ground states at the RK point surrounded by symmetry broken phases~\cite{Rokhsar1988,Leung1996,Syljuasen2005,Syljuasen2006,Syljuasen2004,Moessner2008}. The RK point of these models have been shown to have power-law decaying dimer-dimer and monomer-monomer correlation functions and gapless excitations~\cite{Fisher1961,Kasteleyn1961,Fisher1966a,Fisher1963,Kasteleyn1963a,Sutherland1988b,Sutherland1988a,Kohmoto1988a,Liang1988,Sutherland1988,Kohmoto1988,Levitov1990,Leung1996,Henley1997,Henley1997,Henley2004a,Pollmann2006}. The algebraic decay of monomer correlations is indicative of the logarithmic confinement of electric charges in these models. Thus, such fixed number constrained models on bipartite lattices are expected to have an isolated unstable gapless critical quantum liquid points with logarithmically confined defect and an extensive topological degeneracy. These features of fixed number constrained models are consistent with the effective U(1) gauge theory description of the RK point~\cite{Fradkin2004,Fradkin1990,Moessner2001}; in 2+1D pure U(1) gauge theories are always confining and thus the lack of an extended deconfined topologically ordered phase near the RK point~\cite{PolyakovBook}.

@@

To determine the nature of the ground state at the RK point of the QDPM, we have used Monte Carlo sampling of $\ket{\Psi_{\rm{RK}}}$, using local updates corresponding to the dimer resonances shown in \tabref{tab:pentamer_moves} and single plaquette flips, as defined in \Eqref{eq:HQDM}. Below we present the results of calculations on $L\times L$ square lattices with periodic boundary conditions on systems of linear dimension up to $L=128$.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$. The histograms of these U(1) winding numbers is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$. This confirms that $\Phi_X$ and $\Phi_Y$ are only conserved modulo $3$, rather than strictly conserved as they are in the QDM. We find that the histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors. For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ topological winding numbers $\Phi_X$ and $\Phi_Y$ as measured in for the ground state of the RK point of the QDPM for three different lattice sizes. The colors correspond to calculations initialized to different topological sectors $(\Phi_X,\Phi_Y)$: blue $(0,0)$, red $(1,1)$, and purple $(-1,-1)$. The lines are fits to a Gaussian for each system size. Note that $\Phi$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We will now present a numerical study of the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_3way} (a) we plot the bare dimer-dimer density correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on an $L=32$ lattice; here the origin $\bm{0}$ represents the orange link, and $\bm{r}$ varies over all other links of the lattice. These bare dimer correlations display only short range correlations no obvious symmetry breaking. In \figref{fig:dimer_3way} (b) we show the equal-time dimer-dimer structure factor, 
\begin{equation}
S_d\left( \bm{k} \right) \equiv \frac{1}{ L^2} \sum_{i,j} e^{-i \bm{k} \cdot \left(\bm{r}_i - \bm{r}_j\right)} \left \langle n_{\bm{r}_i} n_{\bm{r}_j} \right \rangle
\end{equation}
where we restrict the sum to horizontal links, for an $L=64$ lattice. 
\figref{fig:dimer_3way} (c) shows $S_d(\bm{k})$ for a particular path through the Brillouin zone. The dimer structure factor displays broad maxima at $\bm{k} = (0, \pm \pi)$ and $\bm{k} = (\pm \pi, \pm \pi)$. The lack of rotational symmetry of $S_d(\bm{k})$ is due to the fact that we have isolated horizontal dimers; the correlations between vertical dimers display the same behavior, rotated by $\pi/2$.
We have plotted the finite size scaling of the intensity of the broad maximum of $S_d$ in  \figref{fig:bragg_scaling}, which we find to vanish in the thermodynamic limit; these maxima are therefore consistent with finite size effects due to short distance correlations. Thus we see no evidence of symmetry breaking in the thermodynamic limit.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_3way.pdf}
        \caption
        {
            (a) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}}
            n_{\bm{r}}\rangle$ on an $L=32$ lattice. The origin corresponds to the orange link, and
            $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off
            the given scale. 
            (b) The horizontal dimer density structure factor $S_d(\bm{k})$ for the
            QDPM RK ground state on an $L=64$ lattice. The lack of $x\leftrightarrow y$ rotational
            symmetry is due to the fact that horizontal links have been isolated. 
            (c) The horizontal dimer-dimer structure factor $S_d(\bm{k})$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:dimer_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\columnwidth]{peak_scaling.pdf}
    \caption{Finite-size scaling of the maxima in the dimer ($S_d$) and magnetic string ($S_M$) structure factors. All maximum vanish in the thermodynamic limit, indicating that the RK state of the QDPM is a symmetric dimer liquid. }
    \label{fig:bragg_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for an $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlations length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_parallel.pdf}
    \caption{The dimer-dimer density correlation function $C_d$ between parallel links in the QDPM RK ground state on a $L=64$ lattice, in the direction perpendicular to the dimer orientation.  The lines represent the best fit to an exponential decay with correlation length $\chi\approx 1$.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the $Z_3$ gauge theory description of the QDPM RK ground state, we study the $Z_3$ magnetic string correlation function as defined in \eqref{eq:Sgamma} \& \eqref{eq:SQDPM}:
\begin{equation}
C_M \left( \bm{r}_{pp'} \right) \equiv \left \langle \exp \left[ \frac{2 \pi}{3} i \left( \Phi_{\gamma_{pp'}} -\tilde{\Phi}_{\gamma_{pp'}} \right) \right] \right \rangle
\end{equation}
where $p$ and $p'$ are two plaquettes, and $\gamma_{pp'}$ is any open string on the dual lattice
connecting $p$ and $p'$. In \figref{fig:vison_3way} (a), we have plotted $|C_M|$ for an $L=32$ lattice. Note, that while $C_M$ depends on the choice of the background dimerization which fixes the gauge, $|C_M|$ is independent of the background dimerization. We find that$C_M$ only displays short range correlations, and no obvious symmetry breaking.

Additionally, we have computed the magnetic string structure factor
\begin{equation}
S_M\left( \bm{k} \right) \equiv \frac{1}{ L^2} \sum_{p,p'} e^{-i \bm{k} \cdot \bm{r}_{pp'}} C_M\left( \bm{r}_{pp'}\right)
\end{equation}
and plot the magnitude $|S_M|$ in \figref{fig:vison_3way} (b). We have chosen a columnar background dimerization comprising horizontal background dimers stacked in columns along the $y$-axis, as shown in \figref{mag_loops_on_orntd_lat.pdf}; this chose breaks the translational symmetry and thus requires doubling the unit cell the $x$-direction. While $|C_M|$ displays no obvious symmetry breaking, there are peaks in $|S_M|$ at $(0,0)$ \& $(0,\pi$) \note{(Owen: What about $(\pi,\pi)$??)}. We have plotted the finite-size scaling of these peaks in \figref{fig:bragg_scaling} which demonstrates that these peaks vanish in the thermodynamic limit.  While the vanishing of the $(0,\pi)$ peak demonstrates the restoration of rotational and translational symmetries in the thermodynamic limit, the vanishing of the $(0,0)$ peak demonstrates the lack of magnetic vortex condensation, which is a mechanism for destroying topological order and breaking lattice symmetries~\cite{Jalabert1991,Ralko2007,Huh2011}. \figref{fig:vison_cor} shows the decay of $|C_M|$ along one of the axes of the dual lattice; we find this decays exponentially with distance, with a correlation length $\chi\approx 2.90$. These exponentially decaying magnetic string correlations are therefore consistent with a symmetric dimer liquid with no magnetic vortex condensation.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{vison_3way.pdf}
        \caption
        {
            (a) Color scale of the magnetic string correlation function 
            on an $L=32$ lattice. The origin corresponds to the orange circle and it has been removed from
            the color scale.
            (b) The magnitude of the $Z_3$ magnetic string correlation function
            $|C_M(\bm{r}_{\bm{0}p})|$, where the correlations is with respect to the orange plaquette.
            (c) The magnitude of the $Z_3$ magnetic string structure factor $|S_M|$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:vison_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_z3_vis.pdf}
    \caption{ The magnitude of the $Z_3$ magnetic string correlation $|C_M|$ along along one axis of the dual lattice of an $L=64$ lattice. The line corresponds to an exponential fit with correlations length \note{$\chi\approx 2.90$}.}
    \label{fig:vison_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To study the confinement of $Z_3$ electrically charged defects in the QDPM, we have Monte Carlo sampled the RK ground state doped with a single pair of monomers. We initialize the system with a monomer on each sub-lattice and supplement the local Monte Carlo updates with updates of the form:
\begin{equation}
\monomermvA  \Longleftrightarrow \monomermvB
\end{equation}
which appear to provide ergodic dynamics in the monomer-dimer-pentamer configuration space. Notice that monomers remain on the same sub-lattice, which is consistent with their assignment of a $Z_3$ electric charge of $\pm1$, where the positive and negative monomers live on different sub-lattices. We then can test for the confinement of monomers by studying the monomer-monomer density correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$, which is plotted in \figref{fig:monomer_correlations}. If the monomers are confined, we expect $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ to decay to zero at large distances; conversely if monomers are deconfined, we expect the monomers to be randomly distributed on the lattice, such that $\langle m_{\bm{0}} m_{\bm{r}} \rangle \sim L^{-4}$ at large separations. \figref{fig:monomer_correlations} demonstrates that $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ decays to $L^{-4}$ for each system size, indicating a deconfinement of monomers.
%%%%%%%%%%%%%%%%%%%%%%%
%\begin{table}
%    \begin{tabular}{| l | c | c |}
%  	\hline
%     & $c$ & $\bar{c}$ \\
%	\hline
%    $i$ & $\monomermvA$ &  $\monomermvB$  \\
%    \hline
%  \end{tabular}
%  \caption{The pair of local dimer \& monomer configurations $c\leftrightarrow\bar{c}$ generate
%  minimal local monomer dynamics. Monomers always remain on their origonal sublattice. }
%  \label{tab:pentamerMoves}
%\end{table}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to $L=64$. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color. \note{Owen: Make the subscripts in the y-axis label bold math. Extend the horizontal lines all the way across the plot. Remove the legend bounding box and use one symbol for each legend entry. Remove the black boundaries of the symbols.}}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state. These results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

@@

A finite gap above the RK point suggests that a stable topologically ordered phase should exist surrounding the RK point. While the Monte Carlo techniques used in this manuscript are limited to the RK point, the phase diagram near the RK point can be explored in future work using quantum Monte Carlo methods, as the QDPM has no sign problem. Away from RK, the condensation of $Z_3$ magnetic vortices could drive a continuous phase transition out of the topologically order phase, analogous to vison condensation in $Z_2$ spin liquids~\cite{Jalabert1991,Ralko2007,Huh2011,Hao2014,Slagle2014}; if such a continuous transition exists in the QDPM it may provide access to a new universality of class of quantum transitions~\cite{XU2012}.

>>>>

The QDM and QFPLM on the square lattice have isolated liquid ground states at the RK point surrounded by symmetry broken phases~\cite{Rokhsar1988,Leung1996,Syljuasen2005,Syljuasen2006,Syljuasen2004,Moessner2008}. The RK point of these models have been shown to have power-law decaying dimer-dimer and monomer-monomer correlation functions and gapless excitations~\cite{Fisher1961,Kasteleyn1961,Fisher1966a,Fisher1963,Kasteleyn1963a,Sutherland1988b,Sutherland1988a,Kohmoto1988a,Liang1988,Sutherland1988,Kohmoto1988,Levitov1990,Leung1996,Henley1997,Henley1997,Henley2004a,Pollmann2006}. The algebraic decay of monomer correlations is indicative of the logarithmic confinement of electric charges in these models. Thus, such fixed number constrained models on bipartite lattices are expected to have an isolated unstable gapless critical quantum liquid points with logarithmically confined defect and an extensive topological degeneracy. These features of fixed number constrained models are consistent with the effective U(1) gauge theory description of the RK point~\cite{Fradkin2004,Fradkin1990,Moessner2001}; in 2+1D pure U(1) gauge theories are confining and thus the lack of an extended deconfined topologically ordered phase near the RK point~\cite{PolyakovBook}.

@@

To determine the nature of the ground state at the RK point of the QDPM, we have Monte Carlo sampled $\ket{\Psi_{\rm{RK}}}$, using local updates corresponding to the dimer resonances shown in \tabref{tab:pentamer_moves} and single plaquette flips, as defined in \Eqref{eq:HQDM}. Below we present the results of calculations on $L\times L$ square lattices with periodic boundary conditions on systems of linear dimension up to $L=128$.

@@

We explicitly demonstrate that the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$, which is shown in \figref{fig:u1_wind_qdpm}. The different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$, and we find that $\Phi_X$ and $\Phi_Y$ are not  strictly conserved (as they are in the QDM), but are only conserved modulo $3$. The histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors. For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ topological winding numbers $\Phi_X$ and $\Phi_Y$ as measured in for the ground state of the RK point of the QDPM for three different lattice sizes. The colors correspond to calculations initialized to different topological sectors $(\Phi_X,\Phi_Y)$: blue $(0,0)$, red $(1,1)$, and purple $(-1,-1)$. The lines are fits to a Gaussian for each system size. Note that $\Phi$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We will now present a numerical study of the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_3way} (a) we plot the bare dimer-dimer density correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on a $L=32$ lattice; here the origin $\bm{0}$ represents the orange link, and $\bm{r}$ varies over all other links of the lattice. These bare dimer correlations display only short range correlations no obvious symmetry breaking. In \figref{fig:dimer_3way} (b) we show the equal-time dimer-dimer structure factor, 
\begin{equation}
S_d\left( \bm{k} \right) \equiv \frac{1}{ L^4} \sum_{i,j} e^{-i \bm{k} \cdot \left(\bm{r}_i - \bm{r}_j\right)} \left \langle n_{\bm{r}_i} n_{\bm{r}_j} \right \rangle
\end{equation}
where we restrict the sum to horizontal links, for a $L=64$ lattice. 
\figref{fig:dimer_3way} (c) shows the magnitude \note{(Owen: Is $S_d$ real \& positive?)} of $S_d(\bm{k})$ for a particular path through the Brillouin zone. The dimer structure factor displays broad maxima at $\bm{k} = (0, \pm \pi)$ and $\bm{k} = (\pm \pi, \pm \pi)$. The lack of rotational symmetry of $S_d(\bm{k})$ is due to the fact that we have isolated horizontal dimers; the correlations between vertical dimers display the same behavior, rotated by $\pi/2$.
We have plotted the finite size scaling of the intensity of the broad maximum of $S_d$ in  \figref{fig:bragg_scaling}, which we find to vanish in the thermodynamic limit; these maxima are therefore consistent with finite size effects due to short distance correlations. Thus we see no evidence of symmetry breaking in the thermodynamic limit.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_3way.pdf}
        \caption
        {
            (a) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}}
            n_{\bm{r}}\rangle$ on an $L=32$ lattice. The origin corresponds to the orange link, and
            $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off
            the given scale. 
            (b) The horizontal dimer density structure factor $S_d(\bm{k})$ for the
            QDPM RK ground state on an $L=64$ lattice. The lack of $x\leftrightarrow y$ rotational
            symmetry is due to the fact that horizontal links have been isolated. 
            (c) The horizontal dimer-dimer structure factor $S_d(\bm{k})$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:dimer_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\columnwidth]{peak_scaling.pdf}
    \caption{Finite-size scaling of the maxima in the dimer ($S_d$) and magnetic string ($S_M$) structure factors. All maxima vanish in the thermodynamic limit, indicating that the RK state of the QDPM is a symmetric dimer liquid. }
    \label{fig:bragg_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlations length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_parallel.pdf}
    \caption{The dimer-dimer density correlation function $C_d$ between parallel links in the QDPM RK ground state on a $L=64$ lattice, along the direction perpendicular to the link orientation.  The lines represent the best fit to an exponential decay with correlation length $\chi\approx 1$.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the $Z_3$ gauge theory description of the QDPM RK ground state, we study the $Z_3$ magnetic string correlation function as defined in \eqref{eq:Sgamma} \& \eqref{eq:SQDPM}:
\begin{equation}
C_M \left( \bm{r}_{pp'} \right) \equiv \left \langle \exp \left[ \frac{2 \pi}{3} i \left( \Phi_{\gamma_{pp'}} -\tilde{\Phi}_{\gamma_{pp'}} \right) \right] \right \rangle
\end{equation}
where $p$ and $p'$ are two plaquettes, $\bm{r}_{pp'}$ is the displacement between $p$ and $p'$,  and $\gamma_{pp'}$ is any open string on the dual lattice
connecting $p$ and $p'$. In \figref{fig:vison_3way} (a), we have plotted $|C_M|$ for a $L=32$ lattice. Note that while $C_M$ depends on the choice of the background dimerization which fixes the gauge, $|C_M|$ is independent of the background dimerization. We find only short range magnetic string correlations correlations that display no obvious symmetry breaking.

Additionally, we have computed the magnetic string structure factor
\begin{equation}
S_M\left( \bm{k} \right) \equiv \frac{1}{ L^4} \sum_{p,p'} e^{-i \bm{k} \cdot \bm{r}_{pp'}} C_M\left( \bm{r}_{pp'}\right)
\end{equation}
and plot the magnitude $|S_M|$ in \figref{fig:vison_3way} (b). We have chosen a columnar background dimerization comprising horizontal background dimers stacked in columns along the $y$-axis, as shown in \figref{fig:orientedSL}; this choice breaks the translational symmetry of the lattice and thus requires doubling the unit cell the $x$-direction. While $|C_M|$ displays no obvious symmetry breaking, there are peaks in $|S_M|$ at $(0,0)$ \& $(0,\pi$). We have plotted the finite-size scaling of these peaks in \figref{fig:bragg_scaling} which demonstrates that these peaks vanish in the thermodynamic limit. We have additionally studied $S_M$ using different background dimerization and find that the location of the finite-size peaks depend on the gauge choice; however in cases we find that these peaks vanish in the thermodynamic limit. The vanishing of the $(0,\pi)$ peak demonstrates the restoration of rotational and translational symmetries in the thermodynamic limit, and the additional vanishing of the $(0,0)$ peak demonstrates the lack of magnetic vortex condensation, which is a mechanism for destroying topological order and breaking lattice symmetries~\cite{Jalabert1991,Ralko2007,Huh2011}. \figref{fig:vison_cor} shows the decay of $|C_M|$ along one of the axes of the dual lattice which we find decays exponentially with distance, with a correlation length $\chi\approx 3$. These exponentially decaying magnetic string correlations are therefore consistent with a symmetric dimer liquid with no magnetic vortex condensation.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{vison_3way.pdf}
        \caption
        {
            (a) Color scale of the $Z_3$ magnetic string correlation function $|C_M(\bm{r}_{\bm{0}p})|$,
            on a $L=32$ lattice where $\bm{0}$ corresponds to plaquette represented by the orange circle.
            (b) The magnitude of the $Z_3$ magnetic string structure factor $S_M$ for a $L=64$ lattice, using a columnar background dimerization as shown in \figref{fig:orientedSL}.             
            (c) $|S_M|$ shown along a path in $k$ space. Statistical errors are smaller than the symbols.
        }
        \label{fig:vison_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_z3_vis.pdf}
    \caption{ The magnitude of the $Z_3$ magnetic string correlation $|C_M|$ along along one axis of the dual lattice of a $L=64$ lattice. The line corresponds to an exponential fit with correlations length $\chi\approx 3$.}
    \label{fig:vison_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To study the confinement of $Z_3$ electrically charged defects in the QDPM, we have Monte Carlo sampled the RK ground state doped with a single pair of monomers. We initialize the system with a monomer on each sub-lattice and supplement the local Monte Carlo updates with updates of the form:
\begin{equation}
\monomermvA  \Longleftrightarrow \monomermvB
\end{equation}
which appear to provide ergodic dynamics in the monomer-dimer-pentamer configuration space. Notice that monomers remain on the same sub-lattice, which is consistent with their assignment of a $Z_3$ electric charge of $\pm1$ (see \tabref{tab:defects}) and the positive and negative monomers live on different sub-lattices. We then can test for the confinement of monomers by studying the monomer-monomer density correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$, which is plotted in \figref{fig:monomer_correlations}. If the monomers are confined, we expect $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ to decay to zero at large distances; conversely if monomers are deconfined, we expect the monomers to be randomly distributed on the lattice, such that $\langle m_{\bm{0}} m_{\bm{r}} \rangle \sim L^{-4}$ at large separations. \figref{fig:monomer_correlations} demonstrates that $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ decays to $L^{-4}$ for each system size, indicating a deconfinement of monomers. This is in contrast with the QDM where monomer correlations decay as a power-law. Thus we find that $Z_3$ electric charges are deconfined in the QDPM.

%%%%%%%%%%%%%%%%%%%%%%%
%\begin{table}
%    \begin{tabular}{| l | c | c |}
%  	\hline
%     & $c$ & $\bar{c}$ \\
%	\hline
%    $i$ & $\monomermvA$ &  $\monomermvB$  \\
%    \hline
%  \end{tabular}
%  \caption{The pair of local dimer \& monomer configurations $c\leftrightarrow\bar{c}$ generate
%  minimal local monomer dynamics. Monomers always remain on their origonal sublattice. }
%  \label{tab:pentamerMoves}
%\end{table}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to $L=64$. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color.}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. Our numerical calculations explicitly show how the extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state. These results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

@@

A finite gap above the RK point suggests that a stable topologically ordered phase should exist surrounding the RK point. While the Monte Carlo techniques used in this manuscript are limited to the RK point, the phase diagram near the RK point can be explored in future work using quantum Monte Carlo methods, as the QDPM has no sign problem. Away from the RK point, the condensation of $Z_3$ magnetic vortices could drive a continuous phase transition out of the topologically order phase, analogous to vison condensation in $Z_2$ spin liquids~\cite{Jalabert1991,Ralko2007,Huh2011,Hao2014,Slagle2014}; if such a continuous transition exists in the QDPM it may provide access to a new universality of class of quantum transitions~\cite{XU2012}.
